---
title: "Capstone Project Write-Up"
subtitle: "Priced Out: Predicting the Next U.S. Housing Crisis"
author:
  - "Brandon Smith"
date: today
bibliography: refs.bib
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
  DATA 510 write-up skeleton (Week 5 Quarto lecture).

  Copy this file to your DS3 repo:
    deliverables/M4-writeup-draft/writeup.qmd

  Week 5: paste prose from your M1 proposal into Introduction, Data, Ethics, Methods.
  Week 7+: replace stubs with real results from notebooks/ and src/.
  Week 12 (M4): methods and findings should be stable; render PDF + HTML for submission.
-->

# Title, Team, and Affiliations

**Project title:** *Priced Out: Predicting the Next U.S. Housing Crisis*

| Name | Role | Willamette email |
|------|------|------------------|
| Brandon Smith | Owner Product Lead | bwsmith2@willamette.edu |

**Peer Stakeholder POs:** *Manish Reddy Kallu | Alexander Jackson | Jackson Garro*

**Repository:** `https://github.com/BSmith-cpu/DATA-510-CAPSTONE-`

# Abstract

*(150 to 250 words after you have results. Summarize: problem, approach, key finding, impact, limitation. Write this section last.)*

# Introduction: Problem and Research Questions

<!-- Paste and revise from M1 proposal Section 2. Note any changes since M1. -->

**Problem statement:**

- Many U.S. metro cities are dealing with the issue of forcing out the middle class. This is a combination of many factors (house values, median income, etc). We are looking at cities already affected by this (Boise, Austin, Tampa) and will use them as training cities to see whether we can identify predictors from that vast data source. 

**Primary research question:**

- Which combination of housing affordability and demographic indicators best predict declines in the share of middle-income households in U.S. metro areas, and how do these predictors appear in Boise City, Idaho, Austin, Texas, and Tampa, Florida between 1975 and 2025 (50 year time span). 

**Secondary questions (if any):**

- TBD

**Changes since M1 proposal:** *(none yet, or explain pivot)*

- None (06/08/2026)
- 

# Impact and Stakeholders

<!-- Paste from M1 proposal Section 3. -->

# Related Work and Background

<!-- Brief cited context. Example: -->

Wildfire smoke and respiratory health are well studied at regional scale [@reid2016; @liu2016]. This project applies county-level public data to Oregon specifically.

# Data and Data Engineering

<!-- Paste from M1 Sections 4 and 5; update with actual pipeline state after M2. -->

**Sources:**

| Source | Role | Access |
|--------|------|--------|
| | | |

**Engineering summary:** Raw data in `data/raw/`; processed panel in `data/processed/`; ingest scripts in `src/ingest/`.

```{r}
#| label: tbl-data-qc
#| tbl-cap: "Example QC summary (replace with your panel)"
#| echo: true

# Replace with your project data when ready:
# panel <- readRDS(here::here("data/processed/analysis_panel.rds"))
# nrow(panel)
knitr::kable(data.frame(
  check = c("Rows", "Counties", "Date range"),
  status = c("stub", "stub", "stub")
))
```

# Ethics and Responsible Use

<!-- Paste from M1 Section 8; expand with project-specific limits. -->

- **Privacy:**
- **Fairness / equity:**
- **Limitations of setting:**
- **Non-causal framing:** *(if observational)*

# Methods and Evaluation

<!-- Paste from M1 Section 6; tighten once analysis is fixed. -->

**Design:**

**Primary analysis:**

**Evaluation metrics:**

**Train / validation / holdout:**

```{r}
#| label: fig-methods-placeholder
#| fig-cap: "Replace with a methods diagram or exploratory plot"
#| echo: false
#| fig-width: 5
#| fig-height: 3

plot(1, 1, type = "n", xlab = "", ylab = "", main = "Methods figure placeholder")
text(1, 1, "Replace after M2/M3")
```

# Results and Visualizations

<!-- Fill after M2 data summary and M3 poster figures. -->

*(Stub: key findings with @fig-methods-placeholder and tables.)*

# Discussion, Limitations, and Future Work

**Interpretation:**

**Limitations:** *(tie to actual data and design)*

**Future work:**

# Conclusions and Recommendations

*(Short, stakeholder-facing takeaways.)*

# Reproducibility and References

**Reproduce main results:**

```bash
cd your-repo
quarto render deliverables/M4-writeup-draft/writeup.qmd
```

**Dependencies:** `requirements.txt` or `renv.lock` in repo root.

**Data:** URLs and approval status documented in `docs/data-dictionary.md`.

# References

::: {#refs}
:::

<!-- Create refs.bib alongside this file. Example entries: -->

<!--
@article{liu2016,
  title={Wildfire-specific fine particulate matter and risk of hospital admissions in urban and rural counties},
  author={Liu, J. C. and others},
  journal={Epidemiology},
  year={2016}
}

@article{reid2016,
  title={Critical review of health impacts of wildfire smoke exposure},
  author={Reid, C. E. and others},
  journal={Environmental Health Perspectives},
  year={2016}
}
-->
