---
title: "Capstone Project Write-Up"
subtitle: "Priced Out: Predicting the Next U.S. Housing Crisis"
author:
  - "Brandon Smith"
date: today
bibliography: refs.bib
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
  DATA 510 write-up skeleton (Week 5 Quarto lecture).

  Copy this file to your DS3 repo:
    deliverables/M4-writeup-draft/writeup.qmd

  Week 5: paste prose from your M1 proposal into Introduction, Data, Ethics, Methods.
  Week 7+: replace stubs with real results from notebooks/ and src/.
  Week 12 (M4): methods and findings should be stable; render PDF + HTML for submission.
-->

# Title, Team, and Affiliations

**Project title:** *Priced Out: Predicting the Next U.S. Housing Crisis*

| Name | Role | Willamette email |
|------|------|------------------|
| Brandon Smith | Owner Product Lead | bwsmith2@willamette.edu |

**Peer Stakeholder POs:** *Manish Reddy Kallu | Alexander Jackson | Jackson Garro*

**Repository:** `https://github.com/BSmith-cpu/DATA-510-CAPSTONE-`

# Abstract

*(150 to 250 words after you have results. Summarize: problem, approach, key finding, impact, limitation. Write this section last.)*

# Introduction: Problem and Research Questions

<!-- Paste and revise from M1 proposal Section 2. Note any changes since M1. -->

**Problem statement:**

Many U.S. metro cities are dealing with the issue of forcing out the middle class. This is a combination of many factors (house values, median income, etc). We are looking at cities already affected by this (Boise, Austin, Tampa) and will use them as training cities to see whether we can identify predictors from that vast data source. 

**Primary research question:**

Which combination of housing affordability and demographic indicators best predict declines in the share of middle-income households in U.S. metro areas, and how do these predictors appear in Boise City, Idaho, Austin, Texas, and Tampa, Florida between 1975 and 2025 (50 year time span). 

**Secondary questions (if any):**

- Do population velocity/accleration signals precede price-to-income ratio spikes, or do they move concurrently?
- Is there a common lag structure (e.g., 1-2 years) between rising housing supply pressure and measureable middle-income displacement across all three training metros?
- Do the same warning signals transfer to metros that have *not* yet collapsed or are Boise/Austin/Tampa idiosyncratic (tech-driven, remote-work-driven, retiree-driven)?

**Changes since M1 proposal:** *(none yet, or explain pivot)*

- None (06/08/2026)
- Removed FRED using the website to create single Ad Hoc for the most at risk cities (07/12/2026)

# Impact and Stakeholders

<!-- Paste from M1 proposal Section 3. -->

# Related Work and Background

Housing affordability research has largely focused on already-unaffordable costal metros (San Francisco, New York, Los Angeles), leaving a gap in predictive work on mid-sized metros *before* they tip [@zillowresearch; @urbaninstitute]. Zillow Research and the Federal Reserve Bank publish retrospective affordability indices, but neither maintains a forward-looking, machine-learning based early-warning system at the CBSA level.

# Data and Data Engineering

<!-- Paste from M1 Sections 4 and 5; update with actual pipeline state after M2. -->

**Sources:**

| Source | Role | Access |
|--------|------|--------|
| Zillow Research (ZHVI, `Metro_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv`) | Median home values by metro, monthly, 2000–2025 | Free, downloaded CSV |
| FHFA House Price Index (`hpi_at_metro.csv`, `hpi_po_metro.csv`) | Quarterly price appreciation index by metro, All-Transactions and Purchase-Only series, 1975–2025 | Free, public FHFA download |
| U.S. Census ACS | Median income, housing cost burden, demographics by CBSA | Free, Census API |
| `merged_full.csv` | R-warehoused join of all sources on CBSA metro code | Internal, generated in `data/final_data/` |

**Engineering summary:** Raw data in `data/raw/`; processed panel in `data/processed/`; ingest scripts in `src/ingest/`.

```{r}
#| label: tbl-data-qc
#| tbl-cap: "Example QC summary (replace with your panel)"
#| echo: true

# Replace with your project data when ready:
# panel <- readRDS(here::here("data/processed/analysis_panel.rds"))
# nrow(panel)
knitr::kable(data.frame(
  check = c("Rows", "Counties", "Date range"),
  status = c("stub", "stub", "stub")
))
```

# Ethics and Responsible Use

<!-- Paste from M1 Section 8; expand with project-specific limits. -->

- **Privacy:**
All data is aggregated at the CBSA metro level (no individual-household-, or address-level records), so no personally identifiable information is included.
- **Fairness / equity:**
Because the model is trained only on Boise, Austin, and Tampa, it may encode feature specific to those metros' growth drivers (tech relocation, retiree migration, remote work) that don't generalize to metros with different ecnomic bases (e.g., manufacturing or college towns). Risk scores must be presetned with this cavet, not as deterministic predictions. 
- **Limitations of setting:**
Displace is under-measured in public data-evictions filings, informal displacement, and household doubling-up are not captured in ACS or FHFA/Zillow series, so the model can only proxy displacement via income-share and affordability metrics, not observe it directly. 
- **Non-causal framing:** *(if observational)*
This is an observational, correlation model. Granger casuality tests in R assess *temporal precedence* of indicators, not casual mechanism - model identifies leading indicators, not policy levers that will fix affordability. 

# Methods and Evaluation

<!-- Paste from M1 Section 6; tighten once analysis is fixed. -->

**Design:**

**Primary analysis:**

**Evaluation metrics:**

**Train / validation / holdout:**

```{r}
#| label: fig-methods-placeholder
#| fig-cap: "Replace with a methods diagram or exploratory plot"
#| echo: false
#| fig-width: 5
#| fig-height: 3

plot(1, 1, type = "n", xlab = "", ylab = "", main = "Methods figure placeholder")
text(1, 1, "Replace after M2/M3")
```

# Results and Visualizations

<!-- Fill after M2 data summary and M3 poster figures. -->

*(Stub: key findings with @fig-methods-placeholder and tables.)*

# Discussion, Limitations, and Future Work

**Interpretation:**

**Limitations:** *(tie to actual data and design)*

**Future work:**

# Conclusions and Recommendations

*(Short, stakeholder-facing takeaways.)*

# Reproducibility and References

**Reproduce main results:**

```bash
cd your-repo
quarto render deliverables/M4-writeup-draft/writeup.qmd
```

**Dependencies:** `requirements.txt` or `renv.lock` in repo root.

**Data:** URLs and approval status documented in `docs/data-dictionary.md`.

# References

::: {#refs}
:::

<!-- Create refs.bib alongside this file. Example entries: -->

<!--
@article{liu2016,
  title={Wildfire-specific fine particulate matter and risk of hospital admissions in urban and rural counties},
  author={Liu, J. C. and others},
  journal={Epidemiology},
  year={2016}
}

@article{reid2016,
  title={Critical review of health impacts of wildfire smoke exposure},
  author={Reid, C. E. and others},
  journal={Environmental Health Perspectives},
  year={2016}
}
-->
