---
title: "Capstone Project Write-Up"
subtitle: "Priced Out: Predicting the Next U.S. Housing Crisis"
author:
  - "Brandon Smith"
date: today
bibliography: refs.bib
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
  DATA 510 write-up skeleton (Week 5 Quarto lecture).

  Copy this file to your DS3 repo:
    deliverables/M4-writeup-draft/writeup.qmd

  Week 5: paste prose from your M1 proposal into Introduction, Data, Ethics, Methods.
  Week 7+: replace stubs with real results from notebooks/ and src/.
  Week 12 (M4): methods and findings should be stable; render PDF + HTML for submission.
-->

# Title, Team, and Affiliations

**Project title:** *Priced Out: Predicting the Next U.S. Housing Crisis*

| Name | Role | Willamette email |
|------|------|------------------|
| Brandon Smith | Owner Product Lead | bwsmith2@willamette.edu |

**Peer Stakeholder POs:** *Manish Reddy Kallu | Alexander Jackson | Jackson Garro*

**Repository:** `https://github.com/BSmith-cpu/DATA-510-CAPSTONE-`

# Abstract

*(150 to 250 words after you have results. Summarize: problem, approach, key finding, impact, limitation. Write this section last.)*

# Introduction: Problem and Research Questions

<!-- Paste and revise from M1 proposal Section 2. Note any changes since M1. -->

**Problem statement:**

Many U.S. metro cities are dealing with the issue of forcing out the middle class. This is a combination of many factors (house values, median income, etc). We are looking at cities already affected by this (Boise, Austin, Tampa) and will use them as training cities to see whether we can identify predictors from that vast data source. 

**Primary research question:**

Which combination of housing affordability and demographic indicators best predict declines in the share of middle-income households in U.S. metro areas, and how do these predictors appear in Boise City, Idaho, Austin, Texas, and Tampa, Florida between 1975 and 2025 (50 year time span). 

**Secondary questions (if any):**

- Do population velocity/accleration signals precede price-to-income ratio spikes, or do they move concurrently?
- Is there a common lag structure (e.g., 1-2 years) between rising housing supply pressure and measureable middle-income displacement across all three training metros?
- Do the same warning signals transfer to metros that have *not* yet collapsed or are Boise/Austin/Tampa idiosyncratic (tech-driven, remote-work-driven, retiree-driven)?

**Changes since M1 proposal:** *(none yet, or explain pivot)*

- None (06/08/2026)
- Removed FRED using the website to create single Ad Hoc for the most at risk cities (07/12/2026)

# Impact and Stakeholders

<!-- Paste from M1 proposal Section 3. -->

# Related Work and Background

Housing affordability research has largely focused on already-unaffordable costal metros (San Francisco, New York, Los Angeles), leaving a gap in predictive work on mid-sized metros *before* they tip [@zillowresearch; @urbaninstitute]. Zillow Research and the Federal Reserve Bank publish retrospective affordability indices, but neither maintains a forward-looking, machine-learning based early-warning system at the CBSA level.

# Data and Data Engineering

<!-- Paste from M1 Sections 4 and 5; update with actual pipeline state after M2. -->

**Sources:**

| Source | Role | Access |
|--------|------|--------|
| Zillow Research (ZHVI, `Metro_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv`) | Median home values by metro, monthly, 2000–2025 | Free, downloaded CSV |
| FHFA House Price Index (`hpi_at_metro.csv`, `hpi_po_metro.csv`) | Quarterly price appreciation index by metro, All-Transactions and Purchase-Only series, 1975–2025 | Free, public FHFA download |
| U.S. Census ACS | Median income, housing cost burden, demographics by CBSA | Free, Census API |
| `merged_full.csv` | R-warehoused join of all sources on CBSA metro code | Internal, generated in `data/final_data/` |

**Engineering summary:** Raw data in `data/raw/`; processed panel in `data/processed/`; ingest scripts in `src/ingest/`.

```{r}
#| label: tbl-data-qc
#| tbl-cap: "Example QC summary (replace with your panel)"
#| echo: true

# Replace with your project data when ready:
# panel <- readRDS(here::here("data/processed/analysis_panel.rds"))
# nrow(panel)
knitr::kable(data.frame(
  check = c("Rows", "Counties", "Date range"),
  status = c("stub", "stub", "stub")
))
```

# Ethics and Responsible Use

<!-- Paste from M1 Section 8; expand with project-specific limits. -->

- **Privacy:**
All data is aggregated at the CBSA metro level (no individual-household-, or address-level records), so no personally identifiable information is included.
- **Fairness / equity:**
Because the model is trained only on Boise, Austin, and Tampa, it may encode feature specific to those metros' growth drivers (tech relocation, retiree migration, remote work) that don't generalize to metros with different ecnomic bases (e.g., manufacturing or college towns). Risk scores must be presetned with this cavet, not as deterministic predictions. 
- **Limitations of setting:**
Displace is under-measured in public data-evictions filings, informal displacement, and household doubling-up are not captured in ACS or FHFA/Zillow series, so the model can only proxy displacement via income-share and affordability metrics, not observe it directly. 
- **Non-causal framing:** *(if observational)*
This is an observational, correlation model. Granger casuality tests in R assess *temporal precedence* of indicators, not casual mechanism - model identifies leading indicators, not policy levers that will fix affordability. 

# Methods and Evaluation

<!-- Paste from M1 Section 6; tighten once analysis is fixed. -->

1. Feature engineering (Python): price-to-income ratio 5-year trend, affordability momentum score (rolling 3-year slope), housing supply pressure (demand growth / permits issued), mortgage burden % of income, population velocity/acceleration, rent-to-own crossover point. 
2. Modeling (Python): XGBoost/LightGBM classifer trained on labled collpase windows from the three training metros, scored against all other metros in the panel. 
3. Explainability (Python): SHAP values to rank features contribution per metro and per time window. 
4. Validation (R): backtesting against known collapse timing, Granger causality tests on top SHAP features, spatial autocorrectilation check (Moran's I) to confirm findings aren't driven by regional clustering (e.g, all SUn Belt metros moving together for unrealted reasons)

**Evaluation metros:** Precision/recall and AUC on held-out collapse windows; lead time as (quarters between flag and observed income-share decline) as the primary stakeholder-facing metric, since flagging accuracy matters less than how much warning time it buys.

**Train / validation / holdout:** Train in Boise, Austin, Tampa (1975-2020 for feature construction, collapse-eriod labels from ~2010-2022 depending on metro); validate in all other metro across the three training cities; holdout test on a small set of currently-affordable mid-sized metros for out-of-sample risk scoring (no ground truth avaibale yet - evaluted qualitatively against domain expectation, e.g., known "hot" metros in housing press). 

```{r}
#| label: fig-methods-placeholder
#| fig-cap: "Replace with a methods diagram or exploratory plot"
#| echo: false
#| fig-width: 5
#| fig-height: 3

plot(1, 1, type = "n", xlab = "", ylab = "", main = "Methods figure placeholder")
text(1, 1, "Replace after M2/M3")
```

# Results and Visualizations

<!-- Fill after M2 data summary and M3 poster figures. -->

*(Stub - populate after M2 EDA and M3 modeling.)* Expected components once real results are in: 

- Time-series comparison of price-to-income ratio for Boise, Austin, Tampa vs. the national median, showing the inflection point before each metro's known collapse. 
- SHAP summary plot ranking the top 5-8 predictive features across the three training metros.
- Ranked risk-index table for the current scoring population of mid-sized metros, with lead-time estimates.
- Backtest table showing precision/recall/AUC and Granger causality p-value for top features.

# Discussion, Limitations, and Future Work

**Interpretation:**

**Limitations:** *(tie to actual data and design)*

- Only three labeled collapse cases, all with somewhat similar drivers (in-migration-driven demand shocks), which may bias the model toward detecing that specific pattern rather than all forms of affordability collapse. 
- FHFA and ACS data lag real-time conditions by months to a year, limiting how "early" any warning can realistically be. 
- No direct displacement or eviction data; middle-income household share is a proxy, not a direct measure. 

**Future work:**

- Expand training labels to additioanl collapse metros (e.g., Nashville, Salt Lake City) to test generalizability beyond the three-city training set.
- Incorporate short-term rental (Airbnb) density and institutional single-family purchase share as additioanl supply-pressure features.
- Extend the R Shiny dashboard to let users adjust risk thresholds interactively for their own stakeholder use case (lender vs. planner vs. advocate).

# Conclusions and Recommendations

*(Short, stakeholder-facing takeaways.)*

# Reproducibility and References

**Reproduce main results:**

```bash
cd your-repo
quarto render deliverables/M4-writeup-draft/writeup.qmd
```

**Dependencies:** `requirements.txt` or `renv.lock` in repo root.

**Data:** URLs and approval status documented in `docs/data-dictionary.md`.

# References

::: {#refs}
:::

<!-- Create refs.bib alongside this file. Example entries: -->

<!--
@misc{zillowresearch,
  title={Zillow Research: Housing Data},
  author={{Zillow Group}},
  year={2026},
  howpublished={\url{https://www.zillow.com/research/data/}}
}

@misc{urbaninstitute,
  title={Housing Finance Policy Center},
  author={{Urban Institute}},
  year={2026},
  howpublished={\url{https://www.urban.org/policy-centers/housing-finance-policy-center}}
}
