---
title: Data Summary 
output: html_document
author: Brandon Smith
---
Affordability Collapse Definition: 
We want to make sure that before moving foward with more feature enginnering that we are defining certain variables and was it considered "unaffordable". 

Condition 1: Price Acceleration - 3-year HPI growth exceeds 30%
Why: This condition captures rapid price increases that may lead to afforadbility issues
'hpi_3yr_growth' >= 0.30

Condition 2: Sustained momentum - YoY ZHVI growth exceeds 10% for 3+ consecutive quarters
Why: This condition caputres sustained price growth that could indicated a consistent trend of rising price, which raises concern for afforadbility.
'zhvi_yoy' >= 0.10 # for 3+ quarters in a row

Condition 3: Affordability burden - median home prices exceeds 5x the median household income
Why: This condition captures the relationship between home prices and income, which is a key factor in afforadbility.
'price_to_income_ratio' >= 5.0

Price Changes Over 1, 3, and 5 Years:
```{r}
merged_full <- read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/merged_full.csv")
```


```{r}
price_changes <- merged_full %>%
  group_by(cbsa) %>%
  arrange(year, qtr) %>%
  mutate(
    # HPI-based price changes (purchase-only, seasonally adjusted)
    hpi_1yr_chg  = index_sa / lag(index_sa, 4)  - 1,  # 4 quarters = 1 year
    hpi_3yr_chg  = index_sa / lag(index_sa, 12) - 1,  # 12 quarters = 3 years
    hpi_5yr_chg  = index_sa / lag(index_sa, 20) - 1,  # 20 quarters = 5 years

    # ZHVI-based price changes (absolute dollar level)
    zhvi_1yr_chg = zhvi_qtr / lag(zhvi_qtr, 4)  - 1,
    zhvi_3yr_chg = zhvi_qtr / lag(zhvi_qtr, 12) - 1,
    zhvi_5yr_chg = zhvi_qtr / lag(zhvi_qtr, 20) - 1
  ) %>%
  ungroup()

glimpse(price_changes)
```


```{r}
price_changes %>%
    filter(str_detect(RegionName, "Austin-Round Rock-San Marcos|Boise City|Tampa")) %>% 
    filter(year >= 2019) %>%
    select(RegionName, year, qtr, hpi_1yr_chg, hpi_3yr_chg, hpi_5yr_chg, zhvi_1yr_chg, zhvi_3yr_chg, zhvi_5yr_chg) %>% 
    arrange(RegionName, year, qtr) %>%
    print(n = 120)
```