---
title: Data Summary 
output: html_document
author: Brandon Smith
---
Affordability Collapse Definition: 
We want to make sure that before moving foward with more feature enginnering that we are defining certain variables and was it considered "unaffordable". 

Condition 1: Price Acceleration - 3-year HPI growth exceeds 30%
Why: This condition captures rapid price increases that may lead to afforadbility issues
'hpi_3yr_growth' >= 0.30

Condition 2: Sustained momentum - YoY ZHVI growth exceeds 10% for 3+ consecutive quarters
Why: This condition caputres sustained price growth that could indicated a consistent trend of rising price, which raises concern for afforadbility.
'zhvi_yoy' >= 0.10 # for 3+ quarters in a row

Condition 3: Affordability burden - median home prices exceeds 5x the median household income
Why: This condition captures the relationship between home prices and income, which is a key factor in afforadbility.
'price_to_income_ratio' >= 5.0

Price Changes Over 1, 3, and 5 Years:
```{r}
library(tidyverse)
library(tidycensus)


merged_full <- read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/merged_full.csv")
```

```{r}
census_api_key("1f51e488b2abfe690f3f5785503381c598f5bb36", install = TRUE, overwrite = TRUE)

years <- 2010:2024
```

```{r}
income_by_year <- map_dfr(years, function(yr) {
  get_acs(
    geography = "cbsa",
    variables = "B19013_001",
    year = yr,
    survey = "acs5"
  ) %>%
  mutate(year = yr)
})
```

```{r}
glimpse(income_by_year)

income_clean <- income_by_year %>%
  transmute(
    cbsa = as.numeric(GEOID),
    year = year,
    median_income = estimate,
    income_moe = moe
  )

write_csv(income_clean, "/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/income_clean.csv")
```

```{r}
glimpse(income_clean)
summary(income_clean$median_income)

# spot-checking for my training cities
income_clean %>%
  filter(cbsa %in% c(12420,14260,45300)) %>%
  print(n = 50)
```


```{r}
merged_full <- merged_full %>%
  left_join(income_clean, by = c("cbsa", "year"))

merged_full <- merged_full %>%
  mutate(
    price_to_income_ratio = zhvi_qtr / median_income
  )

merged_full %>%
  filter(cbsa %in% c(12420, 14260, 45300)) %>%
  select(metro_name.x, year, qtr, zhvi_qtr, median_income, price_to_income_ratio) %>%
  arrange(metro_name.x, year, qtr) %>%
  head(60)
```

```{r}
price_changes <- merged_full %>%
  arrange(cbsa, year, qtr) %>%
  group_by(cbsa) %>%
  mutate(
    # HPI-based price changes
    hpi_1yr_chg  = index_sa / lag(index_sa, 4)  - 1,
    hpi_3yr_chg  = index_sa / lag(index_sa, 12) - 1,
    hpi_5yr_chg  = index_sa / lag(index_sa, 20) - 1,

    # ZHVI-based price changes
    zhvi_1yr_chg = zhvi_qtr / lag(zhvi_qtr, 4)  - 1,
    zhvi_3yr_chg = zhvi_qtr / lag(zhvi_qtr, 12) - 1,
    zhvi_5yr_chg = zhvi_qtr / lag(zhvi_qtr, 20) - 1,

    # flags you were missing
    price_shock = hpi_3yr_chg >= 0.30,
    momentum_flag = !is.na(zhvi_yoy) & zhvi_yoy >= 0.10
  ) %>%
  mutate(
    streak_group = cumsum(lag(!momentum_flag, default = TRUE))
  ) %>%
  group_by(cbsa, streak_group) %>%
  mutate(
    momentum_run = if_else(momentum_flag, row_number(), 0L)
  ) %>%
  ungroup() %>%
  mutate(
    sustained_momentum = momentum_run >= 3,
    collapse_onset_raw_temp = price_shock & sustained_momentum
  ) %>%
  group_by(cbsa) %>%
  mutate(
    collapse_onset_temp = collapse_onset_raw_temp &
      !lag(collapse_onset_raw_temp, default = FALSE),
    collapse_ever_temp = cumany(collapse_onset_raw_temp)
  ) %>%
  ungroup()

glimpse(price_changes)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(year >= 2019) %>%
  select(
    metro_name.x, cbsa, year, qtr,
    hpi_1yr_chg, hpi_3yr_chg, hpi_5yr_chg,
    zhvi_1yr_chg, zhvi_3yr_chg, zhvi_5yr_chg,
    zhvi_yoy, price_shock, momentum_flag, momentum_run,
    sustained_momentum, collapse_onset_temp, collapse_ever_temp
  ) %>%
  arrange(metro_name.x, year, qtr) %>%
  print(n = 100)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(year >= 2019) %>%
  select(
    metro_name.x, year, qtr,
    zhvi_yoy, momentum_flag, momentum_run,
    hpi_3yr_chg, price_shock,
    sustained_momentum, collapse_onset_temp
  ) %>%
  arrange(metro_name.x, year, qtr) %>%
  print(n = 100)
```

```{r}
first_collapse_temp <- price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(collapse_onset_temp) %>%
  group_by(metro_name.x, cbsa) %>%
  summarise(
    first_year = first(year),
    first_qtr = first(qtr),
    hpi_3yr_at_onset = first(hpi_3yr_chg),
    zhvi_yoy_at_onset = first(zhvi_yoy),
    momentum_run_at_onset = first(momentum_run),
    .groups = "drop"
  ) %>%
  arrange(first_year, first_qtr)

print(first_collapse_temp)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(year >= 2019) %>%
  select(
    metro_name.x, year, qtr,
    zhvi_yoy, momentum_flag, momentum_run,
    hpi_3yr_chg, price_shock,
    sustained_momentum, collapse_onset_temp
  ) %>%
  arrange(metro_name.x, year, qtr) %>%
  print(n = 100)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(year >= 2019) %>%
  select(
    metro_name.x, year, qtr,
    zhvi_yoy, momentum_flag, momentum_run,
    hpi_3yr_chg, price_shock,
    sustained_momentum, collapse_onset_temp
  ) %>%
  arrange(metro_name.x, year, qtr) %>%
  print(n = 100)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(price_shock) %>%
  group_by(metro_name.x) %>%
  slice_min(year * 10 + qtr, n = 1) %>%
  select(metro_name.x, year, qtr, hpi_3yr_chg, zhvi_yoy)
  ```

  ```{r}
  price_changes %>%
    filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
    filter(year >= 2018) %>%
    group_by(metro_name.x, year) %>%
    summarise(
      avg_hpi_3yr = mean(hpi_3yr_chCg, na.rm = TRUE),
      avg_zhvi_yoy = mean(zhvi_yoy, na.rm = TRUE), 
      qtrs_price_shock = sum(price_shock, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(metro_name.x, year)
```

```{r}
price_changes %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX|Boise City, ID|Tampa, FL")) %>%
  filter(year >= 2018) %>%
  group_by(metro_name.x, year) %>%
  summarise(
    qtrs_collapse_raw = sum(collapse_onset_raw_temp, na.rm = TRUE),
    any_collapse_onset = any(collapse_onset_temp, na.rm = TRUE),
    ever_collapsed = any(collapse_ever_temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(metro_name.x, year)
```

```{r}
```