---
title: Data Summary 
output: html_document
author: Brandon Smith
---
Affordability Collapse Definition: 
We want to make sure that before moving foward with more feature enginnering that we are defining certain variables and was it considered "unaffordable". 

Condition 1: Price Acceleration - 3-year HPI growth exceeds 30%
Why: This condition captures rapid price increases that may lead to afforadbility issues
'hpi_3yr_growth' >= 0.30

Condition 2: Sustained momentum - YoY ZHVI growth exceeds 10% for 3+ consecutive quarters
Why: This condition caputres sustained price growth that could indicated a consistent trend of rising price, which raises concern for afforadbility.
'zhvi_yoy' >= 0.10 # for 3+ quarters in a row

Condition 3: Affordability burden - median home prices exceeds 5x the median household income
Why: This condition captures the relationship between home prices and income, which is a key factor in afforadbility.
'price_to_income_ratio' >= 5.0

Price Changes Over 1, 3, and 5 Years:

```{r}
library(tidyverse)
library(tidycensus)

merged_full <- read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/merged_full.csv")
```

```{r}
census_api_key("1f51e488b2abfe690f3f5785503381c598f5bb36", install = TRUE, overwrite = TRUE)

years <- 2010:2024
```

```{r}
zhvi_raw <- read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/DATA-510-CAPSTONE--1/data/zhvi_raw.csv")

zhvi_long <- zhvi_raw %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "date",
    values_to = "zhvi"
  ) %>%
  mutate(
    date = as.Date(str_remove(date, "^X") %>% str_replace_all("\\.", "-")),
    year = year(date),
    qtr = quarter(date)
  )

glimpse(zhvi_long)
```

```{r}
zhvi_raw %>%
  filter(str_detect(RegionName, "Boise|Tampa|Austin, TX")) %>%
  select(RegionID, RegionName, RegionType, StateName)
```

```{r}
cbsa_crosswalk <- tribble(
  ~RegionName,       ~cbsa,
  "Austin, TX",      12420,
  "Boise City, ID",  14260,
  "Tampa, FL",       45294
)

zhvi_long_matched <- zhvi_long %>%
  inner_join(cbsa_crosswalk, by = "RegionName")

zhvi_long_matched %>%
  count(cbsa, RegionName)
```

```{r}
zhvi_quarterly <- zhvi_long_matched %>%
  group_by(cbsa, RegionName, year, qtr) %>%
  summarise(zhvi_qtr_fixed = mean(zhvi, na.rm = TRUE), .groups = "drop")

merged_full <- merged_full %>%
  select(-matches("^zhvi_qtr")) %>%
  left_join(zhvi_quarterly, by = c("cbsa", "year", "qtr")) %>%
  rename(zhvi_qtr = zhvi_qtr_fixed)
```

```{r}
merged_full %>%
  filter(str_detect(metro_name.x, "Tampa")) %>%
  distinct(cbsa, metro_name.x)
```

```{r}
training_cbsas <- c(
  austin = 12420,
  boise  = 14260,
  tampa  = 45294
)

write_csv(
  merged_full,
  "/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/merged_full_fixed.csv"
)
```

```{r}
income_by_year <- map_dfr(years, function(yr) {
  get_acs(
    geography = "cbsa",
    variables = "B19013_001",
    year = yr,
    survey = "acs5"
  ) %>%
    mutate(year = yr)
})
```

```{r}
glimpse(income_by_year)

income_clean <- income_by_year %>%
  transmute(
    cbsa = as.numeric(GEOID),
    year = year,
    median_income = estimate,
    income_moe = moe
  )

write_csv(income_clean, "/Users/brandonsmith/DATA-510-CAPSTONE--2/data/final_data/income_clean.csv")
```

```{r}
income_clean_tampa_fix <- income_clean %>%
  filter(cbsa == 45300) %>%
  mutate(cbsa = 45294)

income_clean <- bind_rows(income_clean, income_clean_tampa_fix)

income_clean %>%
  filter(cbsa %in% c(45294, 45300)) %>%
  arrange(cbsa, year)
```

```{r}
income_clean %>%
  filter(cbsa %in% c(12420, 14260, 45294)) %>%
  print(n = 50)
```

```{r}
merged_full <- merged_full %>%
  select(-matches("median_income")) %>%
  select(-matches("income_moe")) %>%
  left_join(income_clean, by = c("cbsa", "year"))

names(merged_full)[str_detect(names(merged_full), "median_income")]
```

```{r}
merged_full <- merged_full %>%
  mutate(
    price_to_income_ratio = zhvi_qtr / median_income,
    affordability_burden = !is.na(price_to_income_ratio) & price_to_income_ratio >= 5.0
  )
```

```{r}
merged_full %>%
  filter(cbsa == 45294, year >= 2019) %>%
  select(cbsa, year, qtr, zhvi_qtr, median_income, price_to_income_ratio, affordability_burden) %>%
  arrange(year, qtr) %>%
  head(30)
```

```{r}
merged_full %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2010) %>%
  select(metro_name.x, cbsa, year, qtr, zhvi_qtr, median_income, price_to_income_ratio, affordability_burden) %>%
  arrange(cbsa, year, qtr) %>%
  head(60)
```

```{r}
merged_full %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX")) %>%
  distinct(cbsa, metro_name.x)

merged_full %>%
  filter(cbsa == 12420) %>%
  summarise(
    n_rows = n(),
    n_zhvi_present = sum(!is.na(zhvi_qtr)),
    min_year_zhvi = min(year[!is.na(zhvi_qtr)], na.rm = TRUE)
  )
```

```{r}
price_changes <- merged_full %>%
  arrange(cbsa, year, qtr) %>%
  group_by(cbsa) %>%
  mutate(
    hpi_1yr_chg = index_sa / lag(index_sa, 4) - 1,
    hpi_3yr_chg = index_sa / lag(index_sa, 12) - 1,
    hpi_5yr_chg = index_sa / lag(index_sa, 20) - 1,

    zhvi_1yr_chg = zhvi_qtr / lag(zhvi_qtr, 4) - 1,
    zhvi_3yr_chg = zhvi_qtr / lag(zhvi_qtr, 12) - 1,
    zhvi_5yr_chg = zhvi_qtr / lag(zhvi_qtr, 20) - 1,
    zhvi_yoy = zhvi_qtr / lag(zhvi_qtr, 4) - 1,

    price_shock = !is.na(hpi_3yr_chg) & hpi_3yr_chg >= 0.30,
    momentum_flag = !is.na(zhvi_yoy) & zhvi_yoy >= 0.10
  ) %>%
  mutate(
    streak_group = cumsum(lag(!momentum_flag, default = TRUE))
  ) %>%
  group_by(cbsa, streak_group) %>%
  mutate(
    momentum_run = if_else(momentum_flag, row_number(), 0L)
  ) %>%
  ungroup() %>%
  mutate(
    sustained_momentum = momentum_run >= 3,
    collapse_onset_raw = price_shock & sustained_momentum & affordability_burden
  ) %>%
  group_by(cbsa) %>%
  mutate(
    collapse_onset = collapse_onset_raw & !lag(collapse_onset_raw, default = FALSE),
    collapse_ever = cumany(collapse_onset_raw)
  ) %>%
  ungroup()

glimpse(price_changes)
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2019) %>%
  select(cbsa, year, qtr, hpi_3yr_chg, zhvi_yoy, price_to_income_ratio, price_shock, momentum_flag, momentum_run, sustained_momentum, collapse_onset_raw, collapse_onset, collapse_ever) %>%
  arrange(cbsa, year, qtr) %>%
  print(n = 100)
```

```{r}
first_collapse <- price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), collapse_onset) %>%
  group_by(cbsa) %>%
  summarise(
    first_year = first(year),
    first_qtr = first(qtr), 
    hpi_3yr_at_onset = first(hpi_3yr_chg),
    zhvi_yoy_at_onset = first(zhvi_yoy),
    pti_at_onset = first(price_to_income_ratio),
    .groups = "drop"
  ) %>%
  arrange(first_year, first_qtr)

glimpse(first_collapse)
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420,14260, 45294), year >= 2018) %>%
  group_by(cbsa, year) %>%
  summarise(
    avg_hpi_3yr = mean(hpi_3yr_chg, na.rm = TRUE),
    avg_zhvi_yoy = mean(zhvi_yoy, na.rm = TRUE), 
    avg_pti = mean(price_to_income_ratio, na.rm = TRUE),
    qtrs_price_shock = sum(price_shock, na.rm = TRUE),
    qtrs_momentum = sum(momentum_flag, na.rm = TRUE),
    qtrs_burden = sum(affordability_burden, na.rm = TRUE),
    any_collapse_onset = any(collapse_onset, na.rm = TRUE), 
    ever_collapsed = any(collapse_ever, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cbsa, year)
```

