---
title: Data Summary 
output: html_document
author: Brandon Smith
---
Affordability Collapse Definition:
Before moving forward with more feature engineering, we define exactly what counts as "unaffordable." The three conditions below were an early, multi-condition candidate definition explored in this notebook; the definition actually used in the final feature set is the simpler `price_to_income_ratio > 5.0` rule built in the last chunk of this file (see the note above that chunk).

Condition 1: Price acceleration - 3-year HPI growth exceeds 30%
Why: captures rapid price increases that may lead to affordability issues.
`hpi_3yr_growth >= 0.30`

Condition 2: Sustained momentum - YoY ZHVI growth exceeds 10% for 3+ consecutive quarters
Why: captures a sustained trend of rising prices, which raises affordability concerns.
`zhvi_yoy >= 0.10` for 3+ quarters in a row

Condition 3: Affordability burden - median home price exceeds 5x median household income
Why: captures the relationship between home prices and income, a key affordability factor.
`price_to_income_ratio >= 5.0`

Price Changes Over 1, 3, and 5 Years:

```{r}
library(tidyverse)
library(tidycensus)

merged_full <- read.csv("../../data/final_data/merged_full.csv")
```

```{r}
# Reads from the CENSUS_API_KEY environment variable rather than hardcoding a
# key in the notebook. Set it once with Sys.setenv(CENSUS_API_KEY = "...") or
# in your .Renviron -- get a free key at https://api.census.gov/data/key_signup.html
census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE, overwrite = TRUE)

years <- 2010:2024
```

```{r}
zhvi_raw <- read.csv("../../data/zhvi_raw.csv")

zhvi_long <- zhvi_raw %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "date",
    values_to = "zhvi"
  ) %>%
  mutate(
    date = as.Date(str_remove(date, "^X") %>% str_replace_all("\\.", "-")),
    year = year(date),
    qtr = quarter(date)
  )

glimpse(zhvi_long)
```

```{r}
zhvi_raw %>%
  filter(str_detect(RegionName, "Boise|Tampa|Austin, TX")) %>%
  select(RegionID, RegionName, RegionType, StateName)
```

```{r}
cbsa_crosswalk <- tribble(
  ~RegionName,       ~cbsa,
  "Austin, TX",      12420,
  "Boise City, ID",  14260,
  "Tampa, FL",       45294
)

zhvi_long_matched <- zhvi_long %>%
  inner_join(cbsa_crosswalk, by = "RegionName")

zhvi_long_matched %>%
  count(cbsa, RegionName)
```

```{r}
zhvi_quarterly <- zhvi_long_matched %>%
  group_by(cbsa, RegionName, year, qtr) %>%
  summarise(zhvi_qtr_fixed = mean(zhvi, na.rm = TRUE), .groups = "drop")

# Only the 3 training cities need this corrected zhvi_qtr (the original
# name-matched zhvi_qtr from merged_full.csv had a known issue for these
# specific metros). Coalescing onto the existing column -- instead of
# dropping it and left-joining a table that only has 3 CBSA codes -- keeps
# the original, broader-coverage zhvi_qtr intact for every other metro.
merged_full <- merged_full %>%
  left_join(zhvi_quarterly, by = c("cbsa", "year", "qtr")) %>%
  mutate(zhvi_qtr = coalesce(zhvi_qtr_fixed, zhvi_qtr)) %>%
  select(-zhvi_qtr_fixed)
```

```{r}
merged_full %>%
  filter(str_detect(metro_name.x, "Tampa")) %>%
  distinct(cbsa, metro_name.x)
```

```{r}
training_cbsas <- c(
  austin = 12420,
  boise  = 14260,
  tampa  = 45294
)

write_csv(
  merged_full,
  "../../data/final_data/merged_full_fixed.csv"
)
```

```{r}
income_by_year <- map_dfr(years, function(yr) {
  get_acs(
    geography = "cbsa",
    variables = "B19013_001",
    year = yr,
    survey = "acs5"
  ) %>%
    mutate(year = yr)
})
```

```{r}
glimpse(income_by_year)

income_clean <- income_by_year %>%
  transmute(
    cbsa = as.numeric(GEOID),
    year = year,
    median_income = estimate,
    income_moe = moe
  )

write_csv(income_clean, "../../data/final_data/income_clean.csv")
```

```{r}
income_clean_tampa_fix <- income_clean %>%
  filter(cbsa == 45300) %>%
  mutate(cbsa = 45294)

income_clean <- bind_rows(income_clean, income_clean_tampa_fix)

income_clean %>%
  filter(cbsa %in% c(45294, 45300)) %>%
  arrange(cbsa, year)
```

```{r}
income_clean %>%
  filter(cbsa %in% c(12420, 14260, 45294)) %>%
  print(n = 50)
```

```{r}
merged_full <- merged_full %>%
  select(-matches("median_income")) %>%
  select(-matches("income_moe")) %>%
  left_join(income_clean, by = c("cbsa", "year"))

names(merged_full)[str_detect(names(merged_full), "median_income")]
```

```{r}
merged_full <- merged_full %>%
  mutate(
    price_to_income_ratio = zhvi_qtr / median_income,
    affordability_burden = !is.na(price_to_income_ratio) & price_to_income_ratio >= 5.0
  )
```

```{r}
merged_full %>%
  filter(cbsa == 45294, year >= 2019) %>%
  select(cbsa, year, qtr, zhvi_qtr, median_income, price_to_income_ratio, affordability_burden) %>%
  arrange(year, qtr) %>%
  head(30)
```

```{r}
merged_full %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2010) %>%
  select(metro_name.x, cbsa, year, qtr, zhvi_qtr, median_income, price_to_income_ratio, affordability_burden) %>%
  arrange(cbsa, year, qtr) %>%
  head(60)
```

```{r}
merged_full %>%
  filter(str_detect(metro_name.x, "Austin-Round Rock-San Marcos, TX")) %>%
  distinct(cbsa, metro_name.x)

merged_full %>%
  filter(cbsa == 12420) %>%
  summarise(
    n_rows = n(),
    n_zhvi_present = sum(!is.na(zhvi_qtr)),
    min_year_zhvi = min(year[!is.na(zhvi_qtr)], na.rm = TRUE)
  )
```

```{r}
price_changes <- merged_full %>%
  arrange(cbsa, year, qtr) %>%
  group_by(cbsa) %>%
  mutate(
    hpi_1yr_chg = index_sa / lag(index_sa, 4) - 1,
    hpi_3yr_chg = index_sa / lag(index_sa, 12) - 1,
    hpi_5yr_chg = index_sa / lag(index_sa, 20) - 1,

    zhvi_1yr_chg = zhvi_qtr / lag(zhvi_qtr, 4) - 1,
    zhvi_3yr_chg = zhvi_qtr / lag(zhvi_qtr, 12) - 1,
    zhvi_5yr_chg = zhvi_qtr / lag(zhvi_qtr, 20) - 1,
    zhvi_yoy = zhvi_qtr / lag(zhvi_qtr, 4) - 1,

    price_shock = !is.na(hpi_3yr_chg) & hpi_3yr_chg >= 0.30,
    momentum_flag = !is.na(zhvi_yoy) & zhvi_yoy >= 0.10
  ) %>%
  mutate(
    streak_group = cumsum(lag(!momentum_flag, default = TRUE))
  ) %>%
  group_by(cbsa, streak_group) %>%
  mutate(
    momentum_run = if_else(momentum_flag, row_number(), 0L)
  ) %>%
  ungroup() %>%
  mutate(
    sustained_momentum = momentum_run >= 3,
    collapse_onset_raw = price_shock & sustained_momentum & affordability_burden
  ) %>%
  group_by(cbsa) %>%
  mutate(
    collapse_onset = collapse_onset_raw & !lag(collapse_onset_raw, default = FALSE),
    collapse_ever = cumany(collapse_onset_raw)
  ) %>%
  ungroup()

glimpse(price_changes)
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2019) %>%
  select(cbsa, year, qtr, hpi_3yr_chg, zhvi_yoy, price_to_income_ratio, price_shock, momentum_flag, momentum_run, sustained_momentum, collapse_onset_raw, collapse_onset, collapse_ever) %>%
  arrange(cbsa, year, qtr) %>%
  print(n = 100)
```

```{r}
first_collapse <- price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), collapse_onset) %>%
  group_by(cbsa) %>%
  summarise(
    first_year = first(year),
    first_qtr = first(qtr), 
    hpi_3yr_at_onset = first(hpi_3yr_chg),
    zhvi_yoy_at_onset = first(zhvi_yoy),
    pti_at_onset = first(price_to_income_ratio),
    .groups = "drop"
  ) %>%
  arrange(first_year, first_qtr)

glimpse(first_collapse)
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420,14260, 45294), year >= 2018) %>%
  group_by(cbsa, year) %>%
  summarise(
    avg_hpi_3yr = mean(hpi_3yr_chg, na.rm = TRUE),
    avg_zhvi_yoy = mean(zhvi_yoy, na.rm = TRUE), 
    avg_pti = mean(price_to_income_ratio, na.rm = TRUE),
    qtrs_price_shock = sum(price_shock, na.rm = TRUE),
    qtrs_momentum = sum(momentum_flag, na.rm = TRUE),
    qtrs_burden = sum(affordability_burden, na.rm = TRUE),
    any_collapse_onset = any(collapse_onset, na.rm = TRUE), 
    ever_collapsed = any(collapse_ever, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cbsa, year)
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2015) %>%
  mutate(
    metro = case_when(
      cbsa == 12420 ~ "Austin",
      cbsa == 14260 ~ "Boise",
      cbsa == 45294 ~ "Tampa"
    ),
    date = as.Date(paste(year, (qtr - 1) * 3 + 1, 1, sep = "-"))
  ) %>%
  ggplot(aes(x = date, y = zhvi_yoy, color = metro)) +
  geom_line(size = 1.5) + 
  geom_hline(yintercept = 5.0, linetype = "dashed", color = "red") + 
  labs(
    title = "Price-to-Income Ratio: Austin, Boise, and Tampa",
    subtitle = "Dashed line = affordability collapse threshold (5.0x)",
    x = NULL,
    y = "Price-to-Income Ratio"
  ) +
  theme_minimal()
```

```{r}
price_changes %>%
  filter(cbsa %in% c(12420, 14260, 45294), year >= 2015) %>%
  mutate(
    metro = case_when(
      cbsa == 12420 ~ "Austin",
      cbsa == 14260 ~ "Boise",
      cbsa == 45294 ~ "Tampa"
    ),
    date = as.Date(paste(year, (qtr - 1) * 3 + 1, 1, sep = "-"))
  ) %>%
  filter(year >= 2015) %>%
  ggplot(aes(x = date, y = metro)) + 
  geom_line(aes(group = metro), color = "grey80", size = 6) + 
  geom_point(
    data = . %>% filter(collapse_onset),
    color = "red",
    size = 4
  ) +
  labs(
    title = "Affordability Collapse Onset by Metro",
    x = NULL,
    y = NULL
  ) + theme_minimal()
```

Final target and feature definition:

This is the pipeline actually used downstream (in `w07-data-summary.ipynb`'s feature engineering and modeling). It supersedes the multi-condition `collapse_onset_raw` definition explored above with a single, simpler rule: a metro is "unaffordable" once its price-to-income ratio exceeds 5.0, and `collapse_onset` marks the first quarter that happens.

**Coverage fix:** this chunk previously re-read a separate `acs_median_income.csv` export that was never committed to the repo, making this notebook impossible to reproduce, and its re-join only reached the 3 training cities. `median_income` is already attached to `merged_full` for every CBSA by the `income_clean` join above (same ACS `B19013_001` pull, all metros, Tampa's `45300` code already remapped to `45294`) -- so this chunk now reuses that column directly instead of depending on a missing file.

  ```{r}
library(tidyverse)

merged_full <- merged_full %>%
  mutate(cbsa = case_when(
    str_detect(metro_name.x, "Austin") ~ 12420,
    str_detect(metro_name.x, "Boise") ~ 14260,
    str_detect(metro_name.x, "Tampa") ~ 45294,
    TRUE ~ cbsa
  ))

merged_full <- merged_full %>%
  arrange(cbsa, year, qtr) %>%
  group_by(cbsa) %>%
  mutate(
    zhvi_yoy = (zhvi_qtr / lag(zhvi_qtr, 4)) - 1,
    price_to_income_ratio = zhvi_qtr / median_income,
    hpi_3yr_chg = (index_sa / lag(index_sa, 12)) - 1,
    is_unaffordable = price_to_income_ratio > 5.0,
    # Some metros (e.g. those with no ACS income match) have price_to_income_ratio
    # -- and therefore is_unaffordable -- entirely NA for the whole group.
    # lag(..., default = FALSE) errors on an all-NA logical group in current
    # dplyr/vctrs; coalesce() after the fact handles it safely instead.
    collapse_onset = coalesce(is_unaffordable, FALSE) & !coalesce(lag(is_unaffordable), FALSE)
  ) %>%
  ungroup()

write_csv(merged_full, "../../data/final_data/price_changes_with_collapse_flags.csv")
  ```