---
title: "EDA on Housing Data"
output: html_document
---

Loading in Basic Libraries
```{r}
# Core data manipulation libraries, visualization libraries, and others
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)
library(janitor)
library(kableExtra)
library(zoo)
```

Loading in Datasets
We are loading in three core datasets that form the foundation of this analysis:
* Zillow Home Value Index (ZHVI)
* Home Price Index (HPI) - Purchase Only
* Home Price Index (HPI) - All Transactions
```{r}
zhvi = read.csv("../../data/zhvi_raw.csv") # nolint

hpi_po <- read.csv(
  "../../data/hpi_po_metro.csv"
)

hpi_at <- read.csv(
  "../../data/hpi_at_metro.csv",
  header    = FALSE,
  col.names = c("metro_name", "cbsa", "yr", "qtr", "index_nsa", "index_sa")
) %>%
  filter(cbsa != "Area Code") %>%
  mutate(
    index_nsa = as.numeric(str_remove_all(as.character(index_nsa), "[()\\s]")),
    index_sa  = as.numeric(str_remove_all(as.character(index_sa),  "[()\\s]"))
  )

glimpse(hpi_po)
glimpse(hpi_at)
```

General Summary of the Datasets
```{r}
glimpse(zhvi)
glimpse(hpi_at)
glimpse(hpi_po)
```
Both the All Transactions and Purchase Only tables look fine; there are some basic cleanup steps still needed, like standardizing headers and removing empty rows.

The ZHVI dataset has an issue: dates are stored as columns, which makes it harder to analyze and to build a proper EDA. We convert it into a long-formatted table below.

Wide to Long Format Conversion

The ZHVI dataset stores each month as its own column, which is not suitable for time series analysis or joining with other datasets. 
We reshape it to long format so each row represents one metro-month observation. 
The leading 'X' and dots in column names are artifacts of how R imports CSV headers.

```{r}
zhvi_long <- zhvi %>%
  pivot_longer(
    cols = starts_with("X"),  # Select all date columns 
    names_to = "date",
    values_to = "zhvi_value"
  ) %>%
  mutate(
    date = str_remove(date, "^X"),  # Remove the leading X
    date = str_replace_all(date, "\\.", "-"),  # Replace dots with hyphens
    date = as.Date(date)  # Convert to Date format
  )
```

We converted the ZHVI dataset from wide to long format, which makes it easier to work with for analysis and visualization.

```{r}
glimpse(zhvi_long)
```

Now that the table looks better, we are going to move ahead to the EDA. 


Quarter Agg of ZHVI

Since HPI data is quarterly, we want to aggregate the ZHVI from monthly to quarterly by taking the mean of the three months within each quarter.
We also want to compute two growth metrics:
* zhvi_yoy - year over year
* zhvi_qoq - quarter over quarter
```{r}
zhvi_qtr <- zhvi_long |>
  mutate(
    yr = year(date),
    qtr = quarter(date)
  ) |>
  group_by(RegionID, RegionName,RegionType, StateName, yr, qtr) |>
  summarise(zhvi_qtr = mean(zhvi_value, na.rm = TRUE), .groups = "drop") |>
  arrange(RegionID, yr, qtr) |>
  group_by(RegionID) |>
  mutate(
    zhvi_yoy = zhvi_qtr / lag(zhvi_qtr, 4) - 1,
    zhvi_qoq = zhvi_qtr / lag(zhvi_qtr, 1) - 1
  ) |>
  ungroup()
```


```{r}
glimpse(zhvi_qtr)
```

Training Cities

We isolate the three training cities - Austin, Boise, and Tampa. The model will learn the pre-collapse signature of these cities.
```{r}
zhvi_train <- zhvi_qtr %>%
  filter(
    str_detect(
      RegionName,
      regex("Austin, TX|Boise City, ID|Tampa, FL", ignore_case = TRUE)
    )
  ) %>%
  mutate(metro_label = RegionName)
```

```{r}
glimpse(zhvi_train)
```

visualization: 
This plot shows the absolute ZHVI level over time for the training cities.
```{r}
# This plot visualizes the quarter over quarter change in House Value from the selected training cities
ggplot(zhvi_train, aes(x = as.Date(as.yearqtr(paste(yr, qtr, sep = " Q"), format = "%Y Q%q")),
                      y = zhvi_qtr,
                      color = RegionName)) +
  geom_line(linewidth = 1.1) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Quarterly Zillow Home Values in Training Metros",
    subtitle = "Austin, Boise, and Tampa",
    x = NULL,
    y = "Average quarterly ZHVI",
    color = NULL
  )
```

This plot shows the rate of change rather than the absolute level. The 2020 - 2022 surge for cities like Boise and Austin having a YoY of 20-30%. 

```{r}
# This plot visualizes the year-over-year (YoY) growth in home values for selected training metros (Austin, Boise, and Tampa) using quarterly Zillow ZHVI data.
ggplot(
  zhvi_train,
  aes(x = as.Date(as.yearqtr(yr + (qtr - 1)/4)),
      y = zhvi_yoy,
      color = RegionName)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Year-over-Year Home Value Growth",
    subtitle = "Training metros using quarterly Zillow ZHVI",
    x = NULL,
    y = "YoY growth",
    color = NULL
  )
```

Cleaning the HPI datasets

We are standardizing the column names and types across HPI files to prepare to join them.
```{r}
hpi_po_clean <- hpi_po%>%
  select(cbsa, metro_name, year = yr, qtr, hpi_po_sa = index_sa)

hpi_at_clean <- hpi_at %>%
  select(cbsa, metro_name, year = yr, qtr, hpi_at_sa = index_sa)
```

Now we are going to be working with CBSA codes to look at specific cities along side the training cities

Loading in Census Population Estimates

We loaded in two Census Bureau CBSA population estimate files, covering 2010-2019 and 2020-2025.
```{r}
pop_2019 <- read_csv("../../data/cbsa-est2019-alldata.csv", show_col_types = FALSE) # nolint
pop_2025 <- read_csv("../../data/cbsa-est2025-alldata.csv", show_col_types = FALSE) # nolint
```

```{r}
glimpse(pop_2019)
glimpse(pop_2025)
```

Cleaning the population data:

The `clean_pop` function filters to Metropolitan Statistical Areas and Metropolitan Divisions (dropping micropolitan areas and counties). The two files are stacked and deduplicated on CBSA + year.

**Coverage note:** FHFA's HPI file reports large multi-division metros (Chicago, New York, Los Angeles, Boston, Detroit, Miami, Philadelphia, San Francisco, Seattle, Washington DC, Tampa, and others) at the *division* level, using the division's own CBSA-like code (e.g. Tampa's `45294` rather than the parent Tampa-St. Petersburg-Clearwater MSA's `45300`). The Census population file stores that division-specific code in a separate `MDIV` column, not `CBSA` -- so the previous version of this function, which only kept `LSAD == "Metropolitan Statistical Area"` rows and used `CBSA`, silently dropped population data for every one of these 37 division-coded metros (confirmed by cross-checking `merged_full.csv`'s CBSA codes against the population files' `MDIV` column). Using `MDIV` as the effective id for division rows recovers all 37.
```{r}
clean_pop <- function(df) {
  df %>%
    filter(LSAD %in% c("Metropolitan Statistical Area", "Metropolitan Division")) %>%
    mutate(cbsa = if_else(LSAD == "Metropolitan Division", MDIV, CBSA)) %>%
    select(cbsa, metro_name = NAME, starts_with("POPESTIMATE")) %>%
    pivot_longer(
      cols      = starts_with("POPESTIMATE"),
      names_to  = "year",
      values_to = "population"
    ) %>%
    mutate(year = as.integer(str_extract(year, "\\d{4}"))) %>%
    filter(!is.na(year))
}

pop_long <- bind_rows(clean_pop(pop_2019), clean_pop(pop_2025)) %>%
  distinct(cbsa, year, .keep_all = TRUE) %>%
  arrange(cbsa, year)

```

```{r}
glimpse(pop_long)
```

Engineering Features: 

From the annual population series, we derive three main features:
* pop_yoy_growth - year-over-year growth rate
* pop_velocity_3yr - 3-year velocity (current population compared to 3 years ago)
* pop_acceleration - change in growth rate

```{r}
pop_features <- pop_long %>%
  group_by(cbsa) %>%
  arrange(year) %>%
  mutate(
    pop_yoy_growth = population /lag(population) -1, # yoy
    pop_velocity_3yr = (population / lag(population, 3)) - 1, # 3 year velocity
    pop_acceleration = pop_yoy_growth - lag(pop_yoy_growth)
  ) %>%
  ungroup()

glimpse(pop_features)
```

CBSA and Zillow crosswalk:

Zillow's files (ZHVI, and ZORI below) use free-text metro names like "New York, NY" rather than CBSA codes, while the Census population file's metro names are the full CBSA title, e.g. "New York-Newark-Jersey City, NY-NJ-PA". A naive join on the full name string only matches metros whose Census name happens to equal Zillow's short name verbatim (single-city, single-state metros) -- multi-city or multi-state metros never match.

We match instead on the first city name (before the first comma **and** before the first hyphen, e.g. "Austin-Round Rock-San Marcos, TX" -> "Austin") **and** the first state abbreviation, since first-city name alone isn't unique across states (e.g. multiple "Columbus" or "Portland" metros). This function is reused for both ZHVI and ZORI below.
```{r}
cbsa_crosswalk <- pop_long %>%
  select(cbsa, census_name = metro_name) %>%
  distinct() %>%
  mutate(
    name_short = str_remove(str_remove(census_name, ",.*$"), "-.*$"),
    state_key  = str_extract(str_remove(census_name, "^[^,]*, "), "^[A-Z]{2}")
  )

# Matches a Zillow long-format table (must have a `RegionName` column, e.g. "New
# York, NY") to a CBSA code via first-city-name + first-state-abbreviation.
match_zillow_to_cbsa <- function(zillow_long_df) {
  lookup <- zillow_long_df %>%
    select(RegionName) %>%
    distinct() %>%
    mutate(
      name_short = str_remove(str_remove(RegionName, ",.*$"), "-.*$"),
      state_key  = str_extract(str_remove(RegionName, "^[^,]*, "), "^[A-Z]{2}")
    ) %>%
    left_join(
      cbsa_crosswalk %>% select(cbsa, name_short, state_key),
      by = c("name_short", "state_key")
    ) %>%
    # A handful of name_short + state_key pairs still map to more than one CBSA
    # (e.g. a metro name that is a prefix of another in the same state); keep
    # the first deterministically rather than silently duplicating rows.
    distinct(RegionName, .keep_all = TRUE)

  zillow_long_df %>%
    left_join(lookup %>% select(RegionName, cbsa), by = "RegionName") %>%
    mutate(cbsa = suppressWarnings(as.integer(cbsa)))
}

zhvi_matched <- match_zillow_to_cbsa(zhvi_long)
message(
  'ZHVI metros matched to a CBSA code: ',
  n_distinct(zhvi_matched$RegionName[!is.na(zhvi_matched$cbsa)]),
  ' / ', n_distinct(zhvi_matched$RegionName)
)
```

```{r}
zhvi_quarterly <- zhvi_qtr %>%
  left_join(zhvi_matched %>% select(RegionName, cbsa) %>% distinct(), by = "RegionName") %>%
  rename(year = yr)
```

Adding a rent-affordability signal (ZORI):

The price-to-income target only measures ownership affordability. Zillow's Observed Rent Index (ZORI) is the rental-market equivalent of ZHVI, published in the same wide monthly format -- we reshape and aggregate it the same way, then match it to CBSA with the same `match_zillow_to_cbsa()` crosswalk used for ZHVI above.
```{r}
zori_raw <- read.csv("../../data/zori_raw.csv")

zori_long <- zori_raw %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "date",
    values_to = "zori_value"
  ) %>%
  mutate(
    date = str_remove(date, "^X"),
    date = str_replace_all(date, "\\.", "-"),
    date = as.Date(date)
  )

zori_matched <- match_zillow_to_cbsa(zori_long)
message(
  'ZORI metros matched to a CBSA code: ',
  n_distinct(zori_matched$RegionName[!is.na(zori_matched$cbsa)]),
  ' / ', n_distinct(zori_matched$RegionName)
)
```

```{r}
zori_qtr <- zori_long %>%
  mutate(yr = year(date), qtr = quarter(date)) %>%
  group_by(RegionName, yr, qtr) %>%
  summarise(zori_qtr = mean(zori_value, na.rm = TRUE), .groups = "drop")

zori_quarterly <- zori_qtr %>%
  left_join(zori_matched %>% select(RegionName, cbsa) %>% distinct(), by = "RegionName") %>%
  rename(year = yr)
```

```{r}
pop_quarterly <- pop_features %>%
  mutate(cbsa = as.integer(cbsa)) %>%
  group_by(cbsa) %>%
  crossing(qtr = as.integer(1:4)) %>%
  ungroup()
```

```{r}
names(pop_quarterly)
glimpse(pop_quarterly)
```
Five-Source Panel Join:

We combine all five data sources into a single quarterly panel dataset joined on CBSA code + year + quarter:
* HPI PO and HPI AT joined with full_join to retain all metro_quarter from either source
* ZHVI quarterly averages attached via left_join
* ZORI (rent) quarterly averages attached via left_join
* Population features broadcast across all four quarters of each year and attached via left_join

A handful of CBSA codes match more than one Zillow `RegionName` (e.g. a short name that's ambiguous even with the state check); `distinct(cbsa, year, qtr, .keep_all = TRUE)` keeps this join one-row-per-metro-quarter rather than silently duplicating rows.
```{r}
hpi_po_clean <- hpi_po %>%
  rename(year = yr) %>%
  select(cbsa, year, qtr, hpi_po_sa = index_sa) %>%
  mutate(cbsa = as.integer(cbsa), year = as.integer(year), qtr = as.integer(qtr))

hpi_at_clean <- hpi_at %>%
  select(cbsa, metro_name, year = yr, qtr, hpi_at_sa = index_sa) %>%
  mutate(cbsa = as.integer(cbsa), year = as.integer(year), qtr = as.integer(qtr))

merged_full <- hpi_at_clean %>%
  full_join(hpi_po_clean, by = c("cbsa", "year", "qtr")) %>%
  left_join(
    zhvi_quarterly %>% distinct(cbsa, year, qtr, .keep_all = TRUE),
    by = c("cbsa", "year", "qtr")
  ) %>%
  left_join(
    zori_quarterly %>% distinct(cbsa, year, qtr, .keep_all = TRUE),
    by = c("cbsa", "year", "qtr")
  ) %>%
  left_join(pop_quarterly,  by = c("cbsa", "year", "qtr")) %>%
  mutate(
    index_sa = coalesce(hpi_at_sa, hpi_po_sa)
  )

glimpse(merged_full)
```

Survivor Analysis:

We are ensuring that we have a sufficiently panel for modeling. So, we are checking how many metros have at least 3 quarters of data across both HPI source and the population series from 2015 forward.
```{r}
problem_rows_at <- hpi_at %>% filter(is.na(suppressWarnings(as.integer(cbsa))) & !is.na(cbsa))
print(unique(problem_rows_at$cbsa))

problem_rows_po <- hpi_po %>% filter(is.na(suppressWarnings(as.integer(cbsa))) & !is.na(cbsa))
print(unique(problem_rows_po$cbsa))
```
Survivor Analysis: 

We are ensuring that we have a sufficiently panel for modeling. So, we are checking how many metros have at least 3 quarters of data across both HPI source and the population series from 2015 forward. 
This window gives us enough data to see what the pre-collapse period was for our training cities and gives candiate metro enough history to compute rolling features. 
We want a minimum of 60 metros. 

```{r}
survival <- merged_full %>%
  filter(year >= 2015) %>%
  group_by(cbsa) %>%
  summarise(
    metro    = first(na.omit(metro_name.x)),
    hpi_obs  = sum(!is.na(index_sa)),
    zhvi_obs = sum(!is.na(zhvi_qtr)),
    pop_obs  = sum(!is.na(population)),
    .groups  = "drop"
  ) %>%
  filter(hpi_obs >= 3, pop_obs >= 3)

cat(sprintf("Metros surviving full join (2015+): %d\n", nrow(survival)))
cat(sprintf("%s 60+ metro target met\n", ifelse(nrow(survival) >= 60, "✅", "❌")))
```

67 metros met the criteria for inclusion in the analysis.

Export merged_full to CSV
```{r}
# Export the merged_full dataframe to CSV
write.csv(merged_full, "../../data/final_data/merged_full.csv", row.names = FALSE)
```