---
title: "EDA on Housing Data"
output: html_document
---

Loading in Basic Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)
library(janitor)
library(kableExtra)
library(zoo)
```

Loading in Datasets
```{r}
zhvi = read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--1/data/Metro_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv") # nolint

hpi_at = read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--1/data/hpi_at_metro.csv") # nolint

hpi_po = read.csv("/Users/brandonsmith/DATA-510-CAPSTONE--1/data/hpi_po_metro.csv") # nolint
```

General Summary of the Datasets
```{r}
glimpse(zhvi)
glimpse(hpi_at)
glimpse(hpi_po)
```
Both the All Transaction and Purchase Only look ok, there are some basic things i might need to go in and change like headers, and removing empty rows

We have an issue when the ZHVI dataset. They decided to put the dates as columns making this a bit harder to analysis and to create a proper EDA. I will be converting this into a long formatted table through R library functions.

Wide to Long Format Conversion
```{r}
zhvi_long <- zhvi %>%
  pivot_longer(
    cols = starts_with("X"),  # Select all date columns (they start with X due to CSV import)
    names_to = "date",
    values_to = "zhvi_value"
  ) %>%
  mutate(
    date = str_remove(date, "^X"),  # Remove the leading X
    date = str_replace_all(date, "\\.", "-"),  # Replace dots with hyphens
    date = as.Date(date)  # Convert to Date format
  )
```

We are converted the ZHVI dataset from wide to long format, which will make it easier to work with for analysis and visualization.

```{r}
glimpse(zhvi_long)
```

Now that the table looks better, we are going to move ahead to the EDA. 

```{r}
zhvi_qtr <- zhvi_long |>
  mutate(
    yr = year(date),
    qtr = quarter(date)
  ) |>
  group_by(RegionID, RegionName,RegionType, StateName, yr, qtr) |>
  summarise(zhvi_qtr = mean(zhvi_value, na.rm = TRUE), .groups = "drop") |>
  arrange(RegionID, yr, qtr) |>
  group_by(RegionID) |>
  mutate(
    zhvi_yoy = zhvi_qtr / lag(zhvi_qtr, 4) - 1,
    zhvi_qoq = zhvi_qtr / lag(zhvi_qtr, 1) - 1
  ) |>
  ungroup()
```


```{r}
glimpse(zhvi_qtr)
```

```{r}
zhvi_train <- zhvi_qtr %>%
  filter(
    str_detect(
      RegionName,
      regex("Austin, TX|Boise City, ID|Tampa, FL", ignore_case = TRUE)
    )
  ) %>%
  mutate(metro_label = RegionName)
```

```{r}
glimpse(zhvi_train)
```

```{r}
# This plot visualizes the quarter over quarter change in House Value from the selected training cities
ggplot(zhvi_train, aes(x = as.Date(as.yearqtr(paste(yr, qtr, sep = " Q"), format = "%Y Q%q")),
                      y = zhvi_qtr,
                      color = RegionName)) +
  geom_line(linewidth = 1.1) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Quarterly Zillow Home Values in Training Metros",
    subtitle = "Austin, Boise, and Tampa",
    x = NULL,
    y = "Average quarterly ZHVI",
    color = NULL
  )
```

```{r}
# This plot visualizes the year-over-year (YoY) growth in home values for selected training metros (Austin, Boise, and Tampa) using quarterly Zillow ZHVI data.
ggplot(
  zhvi_train,
  aes(x = as.Date(as.yearqtr(yr + (qtr - 1)/4)),
      y = zhvi_yoy,
      color = RegionName)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Year-over-Year Home Value Growth",
    subtitle = "Training metros using quarterly Zillow ZHVI",
    x = NULL,
    y = "YoY growth",
    color = NULL
  )
```

Now we are going to be working with CBSA codes to look at specific cities along side the training cities

```{r}
pop_2019 <- read_csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/data/cbsa-est2019-alldata.csv", show_col_types = FALSE) # nolint
pop_2025 <- read_csv("/Users/brandonsmith/DATA-510-CAPSTONE--2/data/cbsa-est2025-alldata.csv", show_col_types = FALSE) # nolint
```

```{r}
glimpse(pop_2019)
glimpse(pop_2025)
```

```{r}
clean_pop <- function(df) {
  df %>%
    filter(LSAD == "Metropolitan Statistical Area") %>%
    select(cbsa = CBSA, metro_name = NAME,
           starts_with("POPESTIMATE")) %>%
    pivot_longer(
      cols      = starts_with("POPESTIMATE"),
      names_to  = "year",
      values_to = "population"
    ) %>%
    mutate(year = as.integer(str_extract(year, "\\d{4}"))) %>% 
    filter(!is.na(year))                                         
}

pop_long <- bind_rows(test_2019, test_2025) %>%
  distinct(cbsa, year, .keep_all = TRUE) %>%
  arrange(cbsa, year)

```

```{r}
glimpse(pop_long)
```

```{r}
pop_features <- pop_long %>%
  group_by(cbsa) %>%
  arrange(year) %>%
  mutate(
    pop_yoy_growth = population /lag(population) -1, # yoy
    pop_velocity_3yr = (population / lag(population, 3)) - 1, # 3 year velocity
    pop_accleration = pop_yoy_growth - lag(pop_yoy_growth)
  ) %>%
  ungroup()

glimpse(pop_features)
```

```{r}
cbsa_crosswalk <- pop_long %>%
  select(cbsa, census_name = metro_name) %>%
  distinct() %>%
  mutate(name_short = str_remove(census_name, ",.*$")) # Remove state information for easier matching

zhvi_cbsa_lookup <- zhvi_long %>%
  select(RegionName) %>%
  distinct() %>%
  mutate(name_short = str_remove(RegionName, ",.*$")) # Remove state information for easier matching

message('ZHVI metros matched to CBSA:', nrow(zhvi_cbsa_lookup)," / ", n_distinct(zhvi_long$RegionName))
```

895/895

```{r}

```

```{r}
```